---
title: "DSC530FinalSMITH"
author: "Champlin Smith"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DSC530 Final Project

# Initial basic model with every feature
```{r}
library(ggplot2)
library(dplyr)
library(car)


data <- read.csv("C:\\Users\\champ\\OneDrive\\Documents\\songs.csv")


model <- lm(track_popularity ~ GenreInt + danceability + energy + key + loudness + speechiness + acousticness + liveness + valence + tempo + duration_ms, data)

summary(model)
```
# QQ-Plot
```{r}
# Q-Q Plot for Track Popularity
qqnorm(data$track_popularity, main="Q-Q Plot of Track Popularity")
qqline(data$track_popularity, col="red")  # Adds a reference line

```
#Scatter plot
```{r}
library(ggplot2)

# Scatter Plot of Popularity vs. Danceability with Genre as Color
ggplot(data, aes(x = danceability, y = track_popularity, color = as.factor(GenreInt))) +
  geom_point(alpha = 0.6) +  # Scatter points with some transparency
  labs(title = "Track Popularity vs. Danceability", 
       x = "Danceability", 
       y = "Track Popularity", 
       color = "Genre") +  # Legend title
  theme_minimal()
```
# Histograms
```{r}
numeric_vars <- names(data)[sapply(data, is.numeric)]

par(mfrow = c(3, 4))
for (var in numeric_vars) {
  hist(data[[var]], main = paste("Histogram of", var), 
       xlab = var, col = "lightblue", border = "black")
}
par(mfrow = c(1, 1))

```
# Descriptive statistics, skewness, kurtosis
```{r}
library(moments)

# Function to compute descriptive statistics
compute_stats <- function(var) {
  var_data <- data[[var]]
  
  mean_val <- mean(var_data, na.rm = TRUE)
  mode_val <- as.numeric(names(sort(table(var_data), decreasing = TRUE)[1]))  # Get most frequent value
  sd_val <- sd(var_data, na.rm = TRUE)
  iqr_val <- IQR(var_data, na.rm = TRUE)
  skew_val <- skewness(var_data, na.rm = TRUE)  # Skewness for tail behavior
  kurt_val <- kurtosis(var_data, na.rm = TRUE)  # Kurtosis for peak and tail thickness
  
  return(data.frame(Variable = var, Mean = mean_val, Mode = mode_val, 
                    SD = sd_val, IQR = iqr_val, Skewness = skew_val, Kurtosis = kurt_val))
}

# Apply to all numeric variables
numeric_vars <- names(data)[sapply(data, is.numeric)]
stats_df <- do.call(rbind, lapply(numeric_vars, compute_stats))

# Print summary statistics
print(stats_df)

```
# PDF Comparisons
```{r}
genre_x <- data %>% filter(GenreInt == 2)
other_genres <- data %>% filter(GenreInt != 2)

compute_pmf <- function(df, variable) {
  pmf <- df %>%
    count(.data[[variable]]) %>%
    mutate(Probability = n / sum(n))
  return(pmf)
}

pmf_genre_x <- compute_pmf(genre_x, "track_popularity")
pmf_other_genres <- compute_pmf(other_genres, "track_popularity")

pmf_genre_x$Scenario <- "EDM"
pmf_other_genres$Scenario <- "Other Genres"

pmf_combined <- bind_rows(pmf_genre_x, pmf_other_genres)

ggplot(pmf_combined, aes(x = track_popularity, y = Probability, fill = Scenario)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  labs(title = "PMF of Track Popularity by Genre",
       x = "Track Popularity",
       y = "Probability") +
  theme_minimal()
```
# CDF
```{r}
ggplot(data, aes(x = track_popularity)) +
  stat_ecdf(geom = "step", color = "blue", size = 1) +
  labs(title = "Cumulative Distribution Function (CDF) of Track Popularity",
       x = "Track Popularity",
       y = "Cumulative Probability") +
  theme_minimal()
```
# Analytic distrbution - histogram vs normal distribution
```{r}
pop_mean <- mean(data$track_popularity, na.rm = TRUE)
pop_sd <- sd(data$track_popularity, na.rm = TRUE)

ggplot(data, aes(x = track_popularity)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = pop_sd), 
                color = "red", size = 1) +
  labs(title = "Histogram of Track Popularity with Normal Distribution",
       x = "Track Popularity",
       y = "Density") +
  theme_minimal()
```
# The distribution of popularity in the dataset is not normal as seen in the above analytic. It is skewed positively and has a kurtosis nearr 5. It has a very heavy tail initially

# Scatter - Danceability
```{r}
ggplot(data, aes(x = danceability, y = track_popularity)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  labs(title = "Scatter Plot: Danceability vs. Track Popularity",
       x = "Danceability",
       y = "Track Popularity") +
  theme_minimal()
```
# Scatter - Energy
```{r}
ggplot(data, aes(x = energy, y = track_popularity)) +
  geom_point(alpha = 0.5, color = "green") +
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  labs(title = "Scatter Plot: Energy vs. Track Popularity",
       x = "Energy",
       y = "Track Popularity") +
  theme_minimal()
```

# Correlation Matrix
```{r}
cor_matrix <- cor(data[, c("danceability", "energy", "track_popularity")], use = "complete.obs", method = "pearson")
print("Correlation")
print(cor_matrix)

cov_matrix <- cov(data[, c("danceability", "energy", "track_popularity")], use = "complete.obs")
print("Covariance")
print(cov_matrix)
```
# The better to understand measure to determine the strength and direction of the relationship between varialbes in Pearsons Correlation, since it provides a set scale from -1 to 1 instead of raw empirical numbers. With this understanding, I do not see any multicollinearity issues as no variables are highly correlated with each other. The highest correlation is -.14, and this is well within bounds to be included in modeling as a feature. 

# Hypothsis Testing. Hypothesis - Energy has no affect on song popularity.
```{r}
model <- lm(track_popularity ~ GenreInt + danceability + energy + key + loudness + speechiness + acousticness + liveness + valence + tempo + duration_ms, data)

summary(model)
```
# Energy is statistically significant, and thus we reject the null hypothesis. Energy has a statistically significant affect on song popularity.


# Step-wise feature selection
```{r}
model <- lm(track_popularity ~ GenreInt + danceability + energy + key + loudness + duration_ms, data)

summary(model)
```

